---
title: "HW03: Exploring `gapminder` using `ggplot2` and `dplyr`"
author: "Emre Erhan"
date: '2018-10-01'
output: github_document
---

## Get libraries
```{r}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(ggplot2))
```

## Task 1: How is life expectancy changing over time on different continents? And which continent has historically had greater life expectancy?

```{r}
gapminder %>%
  select(continent, year, lifeExp) %>%
  arrange(continent, year) %>%
  ggplot(aes(year, lifeExp)) +
  geom_boxplot(aes(group = year)) +
  facet_wrap(~ continent)
```

```{r}
gapminder %>%
  select(continent, year, lifeExp) %>%
  group_by(continent) %>%
  summarise('mean(lifeExp)' = mean(lifeExp), 'sd(lifeExp)' = sd(lifeExp)) %>%
  knitr::kable()
```

The data suggests that Oceania has had the greatest life expectency.

## Task 2: What is the spread of GDP per capita within the continents?


```{r}
gapminder %>%
  ggplot(aes(gdpPercap)) +
  scale_x_log10() +
  geom_density() +
  facet_wrap(~ continent)
```
```{r}
gapminder %>%
  select(continent, gdpPercap) %>%
  group_by(continent) %>%
  summarise('min(gdpPercap)' = min(gdpPercap),
            'max(gdpPercap)' = max(gdpPercap),
            'sd(gdpPercap)' = sd(gdpPercap),
            'mean(gdpPercap)' = mean(gdpPercap)) %>%
  knitr::kable()
```

## Task 3: How did life expectancy and gdp/capita change in Cuba after the 1962 US embargo? 

```{r}
gapminder %>%
  filter(country == "Cuba") %>%
  ggplot(aes(year, gdpPercap)) +
  geom_point(aes(size = lifeExp)) +
  geom_line() +
  geom_vline(aes(xintercept = 1962), color = 'red')
```

Interestingly, life expectancy appears to barely be influenced by the 1962 embargo, while the gdp/capita certainly was. We also see a dip in 1992 that corresponds to the [Cuban Democracy Act](https://en.wikipedia.org/wiki/Cuban_Democracy_Act).

```{r}
gapminder %>%
  filter(country == "Cuba") %>%
  select(year, gdpPercap, lifeExp) %>%
  arrange(year) %>%
  mutate(change_in_gdpPercap = gdpPercap - lag(gdpPercap)) %>%
  knitr::kable()
```

We can see that the greatest change in gdpPercap happened in 1962 and 1992.

## Useful links

I did not need many resources to complete this assignment. These are the ones that I did use:
* Our in-class [cm007](https://github.com/emreerhan/STAT545_participation/blob/master/cm007/cm007-exercise.md) participation document.
* A [StackOverflow link](https://stackoverflow.com/questions/14846547/calculate-difference-between-values-in-consecutive-rows-by-group) describing the `lag` function.